A hazards approach to the biometric analysis of infant mortality

### 2023-03-21 paper2.Rmd
add formula and looking for scania data

### Cohorts by birth season

```{r datain}
library(eha)
Sys.setlocale("LC_ALL", "en_US.utf8")
g <- function(x) (log(x+1))^3 * 365 / (log(366))^3
infa <- readRDS("Rdata/infant.rds")[, c("id", "sex", "birthdate", "enter", "exit", "event")]
infa$exit <- infa$exit / 365 ## NOTE!!
infa$bmonth <- months(infa$birthdate, abbreviate=TRUE)
infa$bmonth <- factor(infa$bmonth, levels = month.abb)
winter <- with(infa, bmonth %in% c("Dec", "Jan", "Feb"))
spring <- with(infa, bmonth %in% c("Mar", "Apr", "May"))
summer <- with(infa, bmonth %in% c("Jun", "Jul", "Aug"))
fall <- with(infa, bmonth %in% c("Sep", "Oct", "Nov"))
infa$bseason <- ""
infa$bseason[winter] <- "winter"
infa$bseason[summer] <- "summer"
infa$bseason[fall] <- "fall"
infa$bseason[spring] <- "spring"
infa$bseason <- factor(infa$bseason, levels = c("spring", "summer", "fall", "winter"))
infa$bdate <- toTime(infa$birthdate)
##infa$bdate <- (infa$bdate - floor(infa$bdate)) * 365
cuts <- seq(0, 1, by = 1 / 12)
comm <- data.frame(month = rep(month.abb, 100))
dat <- make.communal(infa, comm, start = 1861, period = 1 / 12,
                     surv = c("enter", "exit", "event", "bdate"))
dat$month <- factor(dat$month, levels = month.abb)
winter <- with(dat, month %in% c("Dec", "Jan", "Feb"))
spring <- with(dat, month %in% c("Mar", "Apr", "May"))
summer <- with(dat, month %in% c("Jun", "Jul", "Aug"))
fall <- with(dat, month %in% c("Sep", "Oct", "Nov"))
dat$season <- ""
dat$season[winter] <- "winter"
dat$season[spring] <- "spring"
dat$season[summer] <- "summer"
dat$season[fall] <- "fall"
dat$season <- factor(dat$season, levels = c("spring", "summer", "fall", "winter")) 
dat$enter <- 365 * dat$enter # NOTE!
dat$exit <- 365 * dat$exit # NOTE!
```

@@

### Periods by season

@@

Often the Bourheois-Pichat method is applied to tabular data. The most extreme case is
that the only available data are number of births $B$, number of neonatal deaths $D_1$, and total 
number of deaths $D$. The Bourgeois-Pichat line fitting is very simple to do in this case, but 
how about the hazards approach?

It turns out that it is easy to find given the data at hand followíng the procedure given in 
Section 4.3: The probability of dying in the post-neonatal period, given survival of the neonatal
period is estimated by $hat{p} = (D-D_1) / (B - D_1)$, and the length of the post-neonatal time span 
is $g(365) - g(28) \approx 297$ "g-units".

>>>>

### Cohorts by birth month

```{r datain}
library(eha)
nu <- Sys.setlocale("LC_ALL", "en_US.UTF-8")
g <- function(x) (log(x+1))^3 * 365 / (log(366))^3
infa <- readRDS("Rdata/infant.rds")[, c("id", "sex", "birthdate", "enter", "exit", "event")]
infa$exit <- infa$exit / 365 ## NOTE!!
infa$bmonth <- months(infa$birthdate, abbreviate=TRUE)
infa$bmonth <- factor(infa$bmonth, levels = month.abb)
winter <- with(infa, bmonth %in% c("Dec", "Jan", "Feb"))
spring <- with(infa, bmonth %in% c("Mar", "Apr", "May"))
summer <- with(infa, bmonth %in% c("Jun", "Jul", "Aug"))
fall <- with(infa, bmonth %in% c("Sep", "Oct", "Nov"))
infa$bseason <- ""
infa$bseason[winter] <- "winter"
infa$bseason[summer] <- "summer"
infa$bseason[fall] <- "fall"
infa$bseason[spring] <- "spring"
infa$bseason <- factor(infa$bseason, levels = c("spring", "summer", "fall", "winter"))
infa$bdate <- toTime(infa$birthdate)
##infa$bdate <- (infa$bdate - floor(infa$bdate)) * 365
cuts <- seq(0, 1, by = 1 / 12)
comm <- data.frame(month = rep(month.abb, 100))
dat <- make.communal(infa, comm, start = 1861, period = 1 / 12,
                     surv = c("enter", "exit", "event", "bdate"))
dat$month <- factor(dat$month, levels = month.abb)
winter <- with(dat, month %in% c("Dec", "Jan", "Feb"))
spring <- with(dat, month %in% c("Mar", "Apr", "May"))
summer <- with(dat, month %in% c("Jun", "Jul", "Aug"))
fall <- with(dat, month %in% c("Sep", "Oct", "Nov"))
dat$season <- ""
dat$season[winter] <- "winter"
dat$season[spring] <- "spring"
dat$season[summer] <- "summer"
dat$season[fall] <- "fall"
dat$season <- factor(dat$season, levels = c("spring", "summer", "fall", "winter")) 
dat$enter <- 365 * dat$enter # NOTE!
dat$exit <- 365 * dat$exit # NOTE!
```

@@

### Periods by month

@@

Often the Bourgeois-Pichat method is applied to tabular data. The most extreme case is
that the only available data are number of births $B$, number of neonatal deaths $D_1$, and total 
number of deaths $D$. The Bourgeois-Pichat line fitting is very simple to do in this case, but 
how about the hazards approach?

It turns out that it is easy to find given the data at hand followíng the procedure given in 
Section 4.3: The probability of dying in the post-neonatal period, given survival of the neonatal
period is estimated by $\hat{p} = (D-D_1) / (B - D_1)$, and the length of the post-neonatal time span 
is $T = g(365) - g(28) \approx 297$ "g-units". The estimate of the rate $\lambda$ in the post-neonatal period
is then given by

\begin{equation*}
\hat{\lambda} = -\frac{log(1 - \hat{p})}{T}.
\end{equation*}
